{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2023-10-29 paper/si.tex
christian should be senior author!

\def\alld{\texttt{ALLD}}
\def\tft{\texttt{TFT}}
\def\gtft{\texttt{GTFT}}
\def\esm{electronic supplementary material}

@@

\title{{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory}}
\date{}
\author{Nikoleta E. Glynatsi, Alex McAvoy,  Christian Hilbe}

@@

%% BRIEF INTRO %%

\noindent
This document provides further details on our model and our analytical methods, and it contains additional simulation results. 
Section~\ref{section:model} gives a brief summary of our model. 
In particular, we provide further details on our implementation of the evolutionary dynamics, and our use of the rare-mutations limit. 
In Section~\ref{section:analyticalresults} we derive analytical results for the various settings we consider.
These settings differ in what kind of payoff information individuals take into account when updating their strategies. 
In the perfect memory setting, individuals take into account all their interactions against all co-players.
In the limited memory setting, they only consider the very last round of their very last interaction.  
In addition, we describe several model extensions in which the amount of information taken into account is in between these two extremes. 
Finally, Section~\ref{section:furthersimulations} presents further simulation results. 
In particular, we confirm that our main results continue to hold when ({\it i}) players use memory-1 strategies instead of reactive strategies, and  ({\it ii}) mutations are no longer rare.

%%%%%%%%%
%% MODEL  %%
%%%%%%%%%

\section{Description of the model}\label{section:model}

%% MODEL: Summary of the game dynamics %%

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$.
The dynamics unfolds on two time-scales. 
The short time scale describes the game dynamics. 
Here individuals are randomly matched to interact in a repeated prisoner's dilemma.  
Each round individuals can choose whether to cooperate ($C$) or defect ($D$). 
In general, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{equation}\label{eq:donation}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & R & S \\
            \text{defect} & T & P \\
        \end{block}
    \end{blockarray}.
\end{equation}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action, and the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On the short time scale, the players' strategies are fixed.

%% MODEL: Summary of the evolutionary dynamics %%

The long time scale describes the evolutionary dynamics. 
Here, players are allowed to update their strategies based on the payoffs they yield. 
We model these strategy updates with a pairwise comparison process~\citep{traulsen2007pairwise}.
This process assumes that at regular time intervals, one player is randomly selected from the population.
We refer to this player as the `learner' ($L$). 
The learner is then given an opportunity to update its strategy. 
There are two possibilities for how this update may occur. 
With probability $\mu$, the player's strategy mutates randomly. 
In that case, the player's new strategy is drawn uniformly from the space of all reactive strategies~$[0,1]^3$.  
With probability $1\!-\!\mu$, the player engages in a pairwise comparison. 
In that case, the player randomly picks another individual from the population (referred to as the `role model', $R$). 
The learner adopts the role model's strategy with a probability \(\rho\) given by
\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{R}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_{R} - \pi_{L})}}.
\end{equation}
The parameter \(\beta\) is the selection strength
It determines how important payoff differences are for the learner's decision to imitate the role model. 
The parameters $\pi_R$ and $\pi_L$ refer to the relevant payoffs of the role model and the learner, respectively. 
The exact value of these payoffs depend on the players' memory. 
We say players have {\it perfect memory} when $\pi_R$ and $\pi_L$ are given by the players' expected payoffs (across all rounds and across all possible co-players). 
We say players have {\it limited memory} when $\pi_R$ and $\pi_L$ are given by the players' realized payoff in the very last round of their very last interaction partner. 
In addition, we consider several model extensions in which individuals have memory capacities in between these two extreme cases. 
We provide a detailed description of these different cases and the resulting payoffs in Section~\ref{section:analyticalresults}.\\

%% MODEL: Intro to the rare-mutations limit %%

\noindent
{\bf Evolutionary simulations for the rare-mutations limit.} 
To simulate the evolutionary dynamics of the pairwise comparison process described above, it is sometimes useful to assume that mutations are rare, $\mu\!\rightarrow\!0$. 
In that case, whenever a mutant strategy appears, this mutant either fixes in the population or goes extinct before the next mutant appears. 
As a result, at any given time there are at most two different strategies present in the population~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015}. 
This assumptions makes computations more efficient, and it makes some of the results easier to interpret.
In the following, we describe our implementation of the process in the rare-mutations limit in more detail.

%% MODEL: Implementation of the rare-mutations limit %%

Initially, the process starts with a population where all members use the same strategy (called `resident strategy'). 
Then one individual adopts a mutant strategy selected uniformly at random from the set of feasible strategies.
The fixation probability \(\phi_{M}\) of the mutant strategy can be calculated explicitly~\citep{nowak:Nature:2004},
\begin{equation}\label{eq:appendix_fixation_probability}
    \varphi_{M} = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}}.
\end{equation}
Here, the index $k$ corresponds to the current number of players with the mutant strategy (we refer to these players as `mutants'). 
The variables \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of mutants decreases or increases within a single updating step. 
These probabilities depend on the probability that a mutant and a resident are chosen as the learner and the resident, respectively; 
in addition, they depend on the respective switching probability~$\rho$, as described by Eq.~\eqref{Eq:rho}. 
We specify the exact values of  \(\lambda^-_k, \lambda^+_k\) for each memory-setting in the following section.

Depending on the fixation probability \(\varphi_{M}\), the mutant strategy either fixes (becomes the new resident) or goes extinct. 
Afterwards, another random mutant strategy is introduced into the population. 
We iterate this elementary population updating process for a large number of mutant strategies. 
At each step, we record the current resident strategy, and the resulting average cooperation rate. 
We consider this limit of rare-mutations throughout the main text. 
The respective process is summarised by Algorithm~\ref{algorithm:pairwise_comparison}.
However, in Section~\ref{section:furthersimulations} we present additional simulation results to show that our main results continue to hold when the mutation rate is strictly bounded away from zero.

%% MODEL: Algorithm

\begin{algorithm}[t]
  \SetAlgoLined
  $N \leftarrow$ population size\;
  %$k \leftarrow 1$\; 
  resident $\leftarrow$ starting resident\;
   \While{ $t <$ maximum number of steps}
   {mutant $\leftarrow$ random strategy\;
   fixation probability $\leftarrow \varphi_M $\;
   \If{$\varphi_M >$ random: $i \rightarrow [0,1]$}
   {resident $\leftarrow$ mutant;}}
   \caption{Evolutionary process in the limit of rare mutations}\label{algorithm:pairwise_comparison}
\end{algorithm}

\newpage

%%%%%%%%%%%%%%%%%
%% ANALYTICAL RESULTS %%
%%%%%%%%%%%%%%%%%

\section{Analytical results}
\label{section:analyticalresults}

\subsection{Updating Payoffs based on Expected Payoffs (perfect payoff memory)}\label{section:perfect_memory}

Let \(\mathbf{u} = (b-c, -c, b, 0)\)

@@

~\\
{\bf Invasion Analysis}

@@

\subsection{Updating Payoffs based on the Last Round Payoff of One Interaction (limited payoff memory)}\label{section:limited_memory}

@@

\subsection{Updating Payoffs based on the last round payoff of $n$ interactions}\label{section:m_one_n_two}

@@

\subsubsection{Last Round Payoff of Two Interactions}

@@

\subsection{Updating Payoffs based on the Last $m$ Rounds Payoffs of One Interaction}\label{section:m_two_n_one}

@@

\subsection{Updating Payoffs based on the last two rounds payoff of two interactions (\(m=2\) and \(n=2\))}\label{section:m_two_n_two}

@@

%%%%%%%%%%%%%%%%%%
%% FURTHER SIMULATIONS %% 
%%%%%%%%%%%%%%%%%%

\section{Further simulation results}~\label{section:furthersimulations}

\subsection{Simulation Results on the Pairwise Comparison Process}\label{section:simulation_results}

@@

\subsection{Expected and Last Round Updating Payoffs for Memory One Strategies}\label{section:memory_one}

@@

\subsection{Expected and Last Round Updating Payoffs for High Mutation ($\mu \neq 0$)}\label{section:mutation}

In this section we evaluate the main result of this work for \(\mu \neq 0\).
Namely, we explore the evolved population when individuals use perfect and
limited updating payoff memory for different values of \(\mu\). We perform five
independent runs of the pairwise process described in
Section~\ref{section:model}, and at each time step we record the
average player \(\bar{s}=(\bar{y}, \bar{p}, \bar{q})\). The average cooperation
of the resident population for different values of mutation are shown in
Figure~\ref{fig:mutation}. The cooperation rate in the case of perfect
memory is always higher compared to the limited memory regardless of the
mutation value. For mutation value of 1 the processes become random and
this results to a cooperation rate of \(\frac{1}{2}\) in both simulations.

@@

%%%%%%%%%%%%
%% REFERENCES %%
%%%%%%%%%%%%

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography.bib}
}

>>>>

@@

\title{\bf  \sffamily \LARGE Supplementary Information: Evolution of cooperation among individuals with
limited updating payoff memory\\}
\date{}
\author{Nikoleta E. Glynatsi, Alex McAvoy, Christian Hilbe}

@@

Section~\ref{section:pairwise_comparison} gives a brief overview of the pairwise
comparison process. The pairwise comparison process consists of three phases;
(1) the mutation phase (2) the game phase and (3) the update phase. In the
update phase an individual adopts the strategy of another individual based on
their ``updating payoffs''. Here we study different methods for calculating
updating payoffs and we compare their results using as an example the prisoner's
dilemma. In Section~\ref{section:perfect_memory} we describe the conventional
method used in the literature and in Section~\ref{section:limited_memory} we
present a new method which we refer to as 
limited payoff memory. In the limited payoff memory method we assume that
individuals update based on their last round payoff against one other member of
the population. The framework can easily be extended to consider more rounds,
more interactions or both. In
Sections~\ref{section:m_one_n_two}-\ref{section:m_two_n_two} we discuss each of
these extensions and for each extension we present a special case.
In Section~\ref{section:simulation_results} we present numerical results of the
pairwise comparison process using the different updating payoff methods.
In Sections~\ref{section:memory_one} and~\ref{section:mutation} we
verify the main result of this work when we no longer assume (i) that
individuals use reactive strategies but instead memory one strategies (ii) low
mutation.

\section{Pairwise comparison process}\label{section:pairwise_comparison}

Pairwise comparison process is a stochastic process for modelling the evolution
of a finite population. The process starts with assigning all individuals of the
population the same strategy. A strategy is a set of rules of how an individual
should behave in an interaction with another individual. Each elementary time
step of the process consists of three phases; (1) the \textbf{mutation phase}
(2) the \textbf{game phase} and (3) the \textbf{update phase}. These are summarised
in Figure~\ref{fig:pairwise_phases}.

In the \textbf{mutation phase} one individual is
chosen to switch to a new mutant strategy with a probability \(\mu\).
In the \textbf{game phase} individuals are randomly matched with other
individuals in the population and they engage in a repeated game where each
subsequent turn occurs with a fixed probability \(\delta\). At each turn the
individuals decide on an action based on their strategies.
In repeated games there are infinitely many strategies, however, it is commonly
assumed that individuals can only choose strategies from a restricted set. One
such set is that of reactive strategies. A reactive
strategy considers only the previous action of the other player, and thus, a
reactive strategy \(s\) can be written as a three-dimensional vector \(s=(y, p,
q)\). The parameter \(y\) is the probability that the strategy opens with a
cooperation and \(p\), \(q\) are the probabilities that the strategy cooperates
given that the opponent cooperated and defected equivalently.

In the \textbf{update stage} two individuals are randomly selected. From
the two individuals, one  serves as the `learner' and the other as the `role
model'. The learner adopts the role model's strategy with a probability \(\rho\)
given by,

\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{RM}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_{RM}\!-\! \pi_{L})}}.
\end{equation}

\(\pi_{L}\) and \(\pi_{RM}\) are the updating payoffs of the learner and the
role model respectively. The updating payoffs are a measure of how successful
individuals are in the current standing of the population. The parameter
\(\beta\) is known as the selection strength, namely, it shows how important the
payoff difference is when the learner is considering adopting the strategy of
the role model.

\begin{figure}[!htbp]
    \centering
    \includestandalone[width=\textwidth]{static/pairwise_comparison_process_diagram}
    \caption{\textbf{Pairwise comparison process phases.} \textbf{A) Initialisation.} The process begins
    with a finite population where each member is assigned a given strategy.
    Each color represents a different strategy, and the members are labelled by
    letters. \textbf{B) Mutation phase.} An individual is selected (in the
    example individual A) and with a given probability \(\mu\) that individual
    adopts a new strategy. \textbf{C) Game phase.} Individuals are
    selected to interact in a repeated social dilemma with other individuals. We
    demonstrate the case where individuals B and F have been selected to
    interact. They use the reactive strategies \(s_{B} = (y_B, p_B, q_B)\) and
    \(s_{F} = (y_F, p_F, q_F)\) respectively. The opening moves depend on their
    \(y_i\) probability. In turn 1, individual B cooperated,
    thus, F cooperates with a probability \(p_{F}\) in turn 2. On the opposite,
    individual F defected in turn 1, and so B cooperates in the next turn with
    a probability \(q_{B}\). At each turn there is a probability \(\delta\) that a
    subsequent turn will occur, and with a probability (\(1- \delta\)) the interaction ends.
    \textbf{D) Update phase.} At the updating phase two individuals are chosen;
    one serves as the role of the learner and the other one as the 
    role model. In our example C adopts D's strategy with a probability
    \(\rho(\pi_{C}, \pi_{D})\) where \(\pi_{C}, \pi_{D}\) denote the updating
    payoffs of the individuals.}\label{fig:pairwise_phases}
\end{figure}

This elementary step of the process (mutation, game and update phases) is
repeated for a large number of time steps, and at each time step we record the
state of the population.

\subsection{Low mutation \(\mu \rightarrow 0\)}

In the case of low mutation (\(\mu \rightarrow 0\)) we assume that mutations are
rare. In fact, so rare that only two different strategies can be present in the
population at any given time. The case of low mutation is vastly adopted because
it allows us to explicitly calculate the fixation probability of a newly
introduced mutant.

More specifically, the process again starts with a population where all members are of the same
strategy. At each step one individual adopts a mutant strategy randomly selected
from the set of feasible strategies. The fixation probability \(\phi_{M}\) of the
mutant strategy can be calculated explicitly,

\begin{equation}\label{eq:appendix_fixation_probability}
    \varphi_{M} = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}},
\end{equation}

where \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of
mutants decreases and increases respectively, \(N\) is the size of the
population, and \(k\) is the number of mutants. The probabilities \(\lambda^-_k
\text{ and } \lambda^+_k\) depend on the updating payoffs of the mutant and the
resident strategies. Depending on the fixation probability \(\phi_{M}\) the
mutant either fixes (becomes the new resident) or goes extinct. Regardless, in
the elementary time step another mutant strategy is introduced to the
population. We iterate this elementary population updating process for a large
number of mutant strategies and we record the resident strategies at each time
step. The process is summarised by Algorithm~\ref{algorithm:pairwise_comparison}.

\begin{algorithm}[!htbp]
  \SetAlgoLined
  $N \leftarrow$ population size\;
  $k \leftarrow 1$\; resident
   $\leftarrow$ starting resident\;
   \While{ $t <$ maximum number of steps}
   {mutant $\leftarrow$ random strategy\;
   fixation probability $\leftarrow \varphi_M $\;
   \If{$\varphi_M >$ random: $i \rightarrow [0,1]$}
   {resident $\leftarrow$ mutant;}}
   \caption{Evolutionary process}\label{algorithm:pairwise_comparison}
\end{algorithm}

Most of the results we present in this work consider the case of low mutation,
however, we have also verified that the main result holds in the case of
high mutation rates (Section~\ref{section:mutation}).

\subsection{Updating Payoffs}\label{section:updating_payoffs}

The updating payoffs depend on the interactions of the individuals at the game
phase. In this work we assume that in the game phase individuals are matched in
pairs and that they participate in a repeated 2 person donation game. In the
donation game there are two actions: cooperation (\(C\)) and defection (\(D\)).
By cooperating a player provides a benefit \(b\) to the other player at their
cost \(c\), with \(0 < c < b\). Thus the payoffs for a player in each turn are,

\begin{equation}\label{eq:donation}
    \begin{blockarray}{ccc}
        & \text{cooperate} & \text{defect} \\
        \begin{block}{c(cc)}
            \text{cooperate} & b - c & -c \\
            \text{defect} & b & 0 \\
        \end{block}
    \end{blockarray}.
\end{equation}

Let \(\mathbf{u} = (b-c, -c, b, 0)\) be Eq.~\ref{eq:donation} in a vector format, and let \(\mathcal{U}
= \{r, s, t, p\}\) denote the set of feasible payoffs, where \(r\) denotes the
payoff of mutual cooperation, \(s\) the sucker's payoff, \(t\) the temptation to
defect payoff, and \(p\) the punishment payoff.

In the following sections we present several methods for calculating the
updating payoffs. Initially, we discuss the conventional way of the
expected payoffs and afterwards we present our newly introduced methods.

\section{Updating Payoffs based on Expected Payoffs (perfect payoff memory)}\label{section:perfect_memory}

@@

\subsubsection*{Invasion Analysis}

@@

\section{Updating Payoffs based on the Last Round Payoff of One Interaction (limited payoff memory)}\label{section:limited_memory}

@@

\section{Updating Payoffs based on the last round payoff of $n$ interactions}\label{section:m_one_n_two}

@@

\subsection{Last Round Payoff of Two Interactions}

@@

\section{Updating Payoffs based on the Last $m$ Rounds Payoffs of One Interaction}\label{section:m_two_n_one}

@@

\section{Updating Payoffs based on the last two rounds payoff of two interactions (\(m=2\) and \(n=2\))}\label{section:m_two_n_two}

@@

\section{Simulation Results on the Pairwise Comparison Process}\label{section:simulation_results}

@@

\section{Expected and Last Round Updating Payoffs for Memory One Strategies}\label{section:memory_one}

@@

\section{Expected and Last Round Updating Payoffs for High Mutation ($\mu \neq 0$)}\label{section:mutation}

In this section we evaluate the main result of this work for \(\mu \neq 0\).
Namely, we explore the evolved population when individuals use perfect and
limited updating payoff memory for different values of \(\mu\). We perform five
independent runs of the pairwise process described in
Section~\ref{section:pairwise_comparison}, and at each time step we record the
average player \(\bar{s}=(\bar{y}, \bar{p}, \bar{q})\). The average cooperation
of the resident population for different values of mutation are shown in
Figure~\ref{fig:mutation}. The cooperation rate in the case of perfect
memory is always higher compared to the limited memory regardless of the
mutation value. For mutation value of 1 the processes become random and
this results to a cooperation rate of \(\frac{1}{2}\) in both simulations.

@@

\bibliographystyle{unsrt}
\bibliography{bibliography}
